---
title: "Haodong Liu Neuroscience Top 20 Program"
output: html_notebook
---

# Major difference in this field is that dataset "grad" will only include data of Top 10 Neuroscience programs.


```{r}
library(tidyverse)
#library(plotly)
(grad <- read_csv("data/grad.csv",
    col_types = cols_only( 
      uni_name=col_character(),
      major=col_character(),
      degree=col_character(),
      season=col_character(),
      decision=col_character(),
      decision_date=col_character(),
      decision_timestamp=col_double(),
      ugrad_gpa=col_double(),
      gre_verbal=col_double(),
      gre_quant=col_double(),
      gre_writing=col_double(),
      is_new_gre=col_logical(),
      status=col_character())))
problems(grad)
```

```{r}
Top10_Neuro_school = c("Havard University", "Standford University", "University College London", "University Of California, San Francisco (UCSF)","Massachusetts Institute Of Technology (MIT)", "Columbia University","University Of Oxford", "University Of Pennsylvania (UPenn)", "Washington University, St. Louis (WUSTL)", "Johns Hopkins University (JHU)", "University Of California, Los Angeles (UCLA)","University Of California, San Diego (UCSD)", "University Of Cambridge","Yale University","New York University (NYU)","Kings College London","University Of Toronto (UofT)","McGill University","NIH-Karolinska Institutet GPP","NIH Graduate Partnership Program With University Of Karolinska","University Of California, Berkeley (UCB)")

(grad = grad %>% filter( uni_name %in% Top10_Neuro_school == TRUE,
                                     str_detect(major,"Neuroscience") == TRUE))

```

As mentioned previously, we drop variables 'gre_subject','rowid','comments', 'decision_method', 'post_data', and 'post_timestamp'. Aside from this we have no problems regarding data import. While the dataset has some missing data, we keep all data for analyzing variation of single variables.

##Variation of Single Variables:
First we plot counts for the most popular grad schools and programs are.
```{r}
grad %>% group_by(uni_name) %>% count(uni_name) %>% arrange(desc(n))
grad %>% group_by(uni_name,major) %>% count(uni_name, major) %>% arrange(desc(n))

```

From the tables above we see that the most popular college for grad applications is Columbia University with 10,901 applications. For the most popular specific grad programs, we see that Carnegie Mellon University, Computer Science is the most popular with 776 applications. 

Next we assess the selectivity of different grad schools and programs.
```{r}
grad1 <- grad %>% group_by(uni_name) %>%filter(decision == "Accepted") %>% count(uni_name) %>% arrange(desc(n)) 
grad2 <- grad %>% group_by(uni_name) %>% count(uni_name) %>% arrange(desc(n))
colnames(grad1)[2] = "accepted"
merge(grad1,grad2,by =("uni_name")) %>% mutate(rate = accepted/n) %>% arrange(rate) %>% head(10)

grad3 <- grad %>% group_by(uni_name,major) %>% filter(decision == "Accepted") %>% count(uni_name,major) %>% arrange(desc(n)) 
grad4 <- grad %>% group_by(uni_name,major) %>% count(uni_name,major) %>% arrange(desc(n))
colnames(grad3)[3] = "accepted"
merge(grad3,grad4,by=c("uni_name","major")) %>% mutate(rate = accepted/n) %>% arrange(rate) %>% head(10)

merge(grad1,grad2,by =("uni_name")) %>% mutate(rate = accepted/n) %>% filter(uni_name == "Boston University (BU)")
merge(grad3,grad4,by=c("uni_name","major")) %>% mutate(rate = accepted/n) %>% filter(uni_name == "Boston University (BU)") %>% arrange(rate) %>% head(10)

```

Surprisingly, the most selective grad school and grad program are Virginia Consortium with an acceptance rate of 6.25% and University of Colorado, Boulder, Clinical Psychology with an acceptance rate of 1.89% respectively. We also observe that the BU as a whole is not too selective with an acceptance rate of 42.8%. We also obeserve that the most selective major at BU is genetic counseling with only 1 student accepted out of 20 (5%).

Next, we look at the change in number of application over time.
```{r}
# Decision reported over time (2015, 2016, 2017)?
grad$decision_date <- grad$decision_date %>% str_replace("\\(", "") %>% str_replace("\\)", "") 
# Create a dataset for plotting number of application verses year
grad_year = grad %>% select(degree, decision_date)  %>% 
  filter(!is.na(decision_date)) %>%
  mutate(yr = str_match(decision_date, "...\\d$")) %>%
  filter(degree == "MFA" | degree == "MS" | degree == "PhD") %>%
  filter(as.integer(yr) < 2016 ) %>% filter(as.integer(yr) > 2005)
grad_year$decision_date <- NULL
# plot
grad_year %>% group_by(yr, degree) %>% ggplot(aes(x = as.factor(yr), fill = degree)) + geom_bar(position = "dodge")+ 
labs(x ="Year",
     y ="Count",
     title="Number of Application for each Degree Type by Year")
```


The dataset has official data of report from 2006 to 2015. The application report of three degrees, MFA, MS, PhD increase each year until 2015. The overall shape has a plateauing trend.

Next, we look at the distribution of applications by season and the counts of each admission decision.
```{r}
# bar chart for season
# bar chart includes both Spring semester and Fall semester
grad %>% 
  filter(!is.na(season)) %>%
  ggplot() +
  geom_bar(aes(x = season)) +
  labs(title = "Admission Season")
# bar chart for Fall semester only
grad %>%
  filter(!is.na(season)) %>%
  group_by(fall = str_match(season, "^\\F..")) %>%
  filter(!is.na(fall)) %>%
  ggplot() +
  geom_bar(aes(x = season)) +
  labs(title = "Admission Season (Fall)")
# bar chart for Spring semester only
grad %>%
  filter(!is.na(season)) %>%
  group_by(fall = str_match(season, "^\\F..")) %>%
  filter(is.na(fall)) %>%
  ggplot() +
  geom_bar(aes(x = season)) +
  labs(title = "Admission Season (Spring)")
# bar chart for decision
grad %>%
  filter(!is.na(decision)) %>%
  ggplot() +
  geom_bar(aes(x = decision)) +
  labs(title = "Admission Decision")
```

We see that the total number of application are signicantly higher for the fall semester than the Spring semester. Focusing only on fall semesters, graduate school enrollment clearly has shown a positive trend over the years. Spring enrollment does not show an explicit pattern. When looking at the distribution of decisions, the majority of candidates either recieve of report mainly acceptances and rejections, while a few candiates recieve other forms of responses such as waitlist, interview, or "other."

Next, we plot the distribution of GRE test scores, and GPA. Because these is a variable "is_new_gre", which distinguished between old and new GRE, we filter for only new GRE scores, as the majority of observations report new GRE scores.
```{r}
# GRE Verbal
grad %>% select(gre_verbal ,is_new_gre) %>% 
filter(is_new_gre == TRUE & is.na(gre_verbal)!= TRUE ) %>% ggplot + geom_histogram(aes(gre_verbal)) + 
labs(x ="GRE Verbal Score",
     y ="Count",
     title="Frequencies of GRE Verbal Scores")
# GRE quant
grad %>% select(gre_quant ,is_new_gre) %>% 
  filter(is_new_gre == TRUE & is.na(gre_quant)!= TRUE ) %>% ggplot + geom_histogram(aes(gre_quant)) + 
labs(x ="GRE Quant Score",
     y ="Count",
     title="Frequencies of GRE Quant Scores")
# GRE writing
grad %>% select(gre_writing ,is_new_gre) %>% 
  filter(is_new_gre == TRUE & is.na(gre_writing)!= TRUE) %>% ggplot + geom_histogram(aes(gre_writing)) + 
labs(x ="GRE Writing Score",
     y ="Count",
     title="Frequencies of GRE Writing Scores")
```

We see from the above histograms that GRE verbal scores range from 130 to 170 with a bell shape. Most of them concentrate 155 - 165. GRE quant score range from 130 to 170 with step like shape. Scores tend to concentrate 160 - 170. GRE writing scores range from 2 to 6 with a bell like shape. Most people get a score of 4.

```{r}
grad %>% filter(!is.na(ugrad_gpa) & ugrad_gpa < 4.0) %>% 
  ggplot(aes(ugrad_gpa)) + geom_histogram(bins = 40) + labs(titles = "GPA Distribution")
```

We see that the distribution of GPAs for the observations tend to be left skewed, with the majority of candidates having more than 3.6 GPA. This is accepted as grad programs tend to look at GPA as a major factor, and students who aim to attend a grad school would likely have higher GPAs.

Lastly, we look at the distribution of student status (internation, US, international with US degree, etc)
```{r}
grad %>% filter(!is.na(status)) %>% 
  mutate(count = n()) %>% 
  ggplot(aes(x = status)) + geom_bar() + 
  labs(titles = "Immigration Status") 
```

From the chart above, we see that the majority of students applying are American. In Immigration Status, around 60% of applicants are American and the rest of them are international students. We can tell that a big amount of graduate or Ph.D. students are coming from an international background. 

## Covariation Between Multiple Variables
One covariation of interest is the influence of student status (internation, US, etc.) vs admission decision. 
```{r}
# student identity vs acceptance rate
# table for student status vs decision
(grid <- grad %>%
  filter(!is.na(status), !is.na(decision)) %>%
  group_by(status, decision) %>%
  summarise(count = n()) %>%
  spread(key = decision, value = count))
# bar chart 
grad %>%
  filter(!is.na(decision), !is.na(status)) %>%
  ggplot() +
  geom_bar(aes(status, fill = decision), position = "dodge") +
  labs(title = "Admission Decision vs Status")
# acceptance rate based on student status
grid$sum = rowSums(grid[, c("Accepted", "Interview", "Other", "Rejected", "Wait listed")])
grid %>%
  gather('Accepted', 'Interview', 'Other', 'Rejected', 'Wait listed', 
         key = "decision", value = "cases") %>%
  mutate(prop = cases/sum) %>%
  filter(status != "Other") %>%
  ggplot(aes(group = decision, status, prop, color = decision)) + 
  geom_line()
```

From the charts above, it seems that US based students tend to have higher acceptance rates than international students, and internation with US degree students. The bar chart shows that, for American students, the number of getting accepted is higher than the number of getting rejected. However, for international students and international students with US degree, the number of acceptance is lower than the number of rejection. To further investigate if international students are treated differently, we calculate the decision rate. For each status category, we divide the total number of each admission decision by total number of students to get the decision rate. From the plot we can tell that the proportion of getting accepted is higher for American students than international students, and the proportion of getting rejected is higher for international students with US degree.


Another covariation of interest is the relationship between GPA and GREE scores. For this we summed GRE verbal and GRE quant to get the full GRE score, and created a scatter plot against GPA. We filtered GPA to be less than 4, as GPA of different scales are not comparable.
```{r}
grad %>% filter(!is.na(ugrad_gpa|gre_verbal|gre_quant)& ugrad_gpa < 4 & ugrad_gpa >1, is_new_gre == TRUE) %>% mutate(GRE_Total = gre_verbal + gre_quant) %>% group_by(uni_name) %>% mutate(mean_gpa = mean(ugrad_gpa), mean_GRE = mean(GRE_Total)) %>% ungroup() %>%
ggplot(aes(x = mean_GRE, y = mean_gpa)) + geom_point(aes(color = "RED", alpha = 0.001)) +
  labs(titles = "Relationship between GPA and GRE Score",
       y = "GPA",
       x = "GRE Score")
```

From the plot above the relationship between GPA and GRE seems to be positively correlated but is not as strong of a relationship as we expected. Most GPAs tend to be on the higher range: people densely fall into the range between 3.5 and 3.75; GRE scores seem to be more variable across application: scores for all applicants concentrate in the range between 300 and 325 with more outliers. 

Lastly, to measure the correlation across all continuous variables, we create a scatterplot matrix, and correlation matrix.

```{r}
g <- grad[complete.cases(grad),] %>% mutate(acceptance = decision == "Accepted") %>% filter(ugrad_gpa<=4,is_new_gre == TRUE) %>% select(ugrad_gpa, gre_verbal,gre_quant,gre_writing,acceptance) %>% pairs()
```

This plot is somewhat unclear due to the verse dense concentration of the datapoints. In the next step, we will likely use regression to model the probability of acceptance based on the different covariates.